---
title: "Surgical Complication Prediction"
author: "Hiba Khan"
date: "2026-05-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
params:
  data_path: "/share/dsa435601s26/data/"
  my_path:   "/share/dsa435601s26/hakhan4/"
  load_run:  FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6
)

seed <- 2024
set.seed(seed)

.libPaths(c("/share/dsa435601s26/installed_libraries", .libPaths()))

library(dplyr)
library(tidyr)
library(readxl)
library(skimr)
library(janitor)
library(rsample)
library(fastDummies)
library(recipes)
```

------------------------------------------------------------------------

## 1. Read and Explore the Dataset

```{r load-data}
data_file <- paste0(params$data_path, "Surgical-deepnet.csv")
df <- read.csv(data_file, sep = ",") %>%
  janitor::clean_names()

cat("Raw data dimensions:\n")
print(dim(df))
print(names(df))
```

### Outcome Distribution

The target variable `complication` is binary: 1 = complication occurred, 0 = no complication. We expect moderate class imbalance.

```{r outcome-distribution}
outcome_distribution <- df %>%
  count(complication) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

knitr::kable(outcome_distribution, caption = "Distribution of outcome (complication)")

# Confirm binary outcome
stopifnot(all(df$complication %in% c(0, 1)))
```

------------------------------------------------------------------------

## 2. Scope-Aligned Filtering and Outcome Preparation

`mort30` is removed because the objective is to predict surgical complications, not mortality. No patient subgroup filtering is applied in the final scope.

```{r filter-data}
df_filtered <- df %>%
  select(-mort30)

cat("Filtered data dimensions:\n")
print(dim(df_filtered))
```

```{r skim-filtered, results='markup'}
skimr::skim(df_filtered)
```

------------------------------------------------------------------------

## 3. Train / Test Partitioning

A random 70/30 split stratified by `complication` ensures both sets preserve the outcome balance.

```{r split-data}
split    <- initial_split(df_filtered, prop = 0.70, strata = "complication")
trainDat <- training(split)
testDat  <- testing(split)

cat("Training rows:", nrow(trainDat), "\n")
cat("Testing rows: ", nrow(testDat),  "\n")
cat("Split check:  ", nrow(trainDat) + nrow(testDat) == nrow(df_filtered), "\n")
```

------------------------------------------------------------------------

## 4. Missing-Value Checks

```{r missing-values}
missing_summary <- sapply(trainDat, function(x) sum(is.na(x)))
missing_summary <- missing_summary[missing_summary > 0]

if (length(missing_summary) == 0) {
  cat("No missing values found. Imputation is not required.\n")
} else {
  print(missing_summary)
}
```

------------------------------------------------------------------------

## 5. One-Hot Encoding

Dummy variables are created for `asa_status`, `ahrq_ccs`, and `race`. Original columns are retained alongside the dummies. Dummy columns are aligned across train and test to handle categories that appear in only one split.

```{r dummy-encoding}
categorical_to_dummy <- c("asa_status", "ahrq_ccs", "race")

trainDat <- fastDummies::dummy_cols(
  trainDat,
  select_columns        = categorical_to_dummy,
  remove_selected_columns = FALSE,
  remove_first_dummy    = FALSE
)

testDat <- fastDummies::dummy_cols(
  testDat,
  select_columns        = categorical_to_dummy,
  remove_selected_columns = FALSE,
  remove_first_dummy    = FALSE
)

# Align columns across splits
in_train_not_test <- setdiff(names(trainDat), names(testDat))
in_test_not_train <- setdiff(names(testDat),  names(trainDat))

for (col in in_train_not_test) testDat[[col]]  <- 0
for (col in in_test_not_train)  trainDat[[col]] <- 0

testDat <- testDat[, names(trainDat)]
stopifnot(isTRUE(all.equal(names(trainDat), names(testDat))))

cat("Dummy encoding complete. Columns in training set:", ncol(trainDat), "\n")
```

------------------------------------------------------------------------

## 6. Standardization

Continuous predictors are standardized using training-set means and standard deviations. The same parameters are applied to the test set to prevent data leakage.

```{r standardize}
standardize_features <- function(data, features, suffix = "_std", params = NULL) {
  computed_params <- list()

  for (feat in features) {
    if (!feat %in% names(data)) {
      warning("Skipping missing feature: ", feat)
      next
    }

    if (is.null(params)) {
      feat_mean <- mean(data[[feat]], na.rm = TRUE)
      feat_sd   <- sd(data[[feat]],   na.rm = TRUE)
    } else {
      feat_mean <- params[[feat]]$mean
      feat_sd   <- params[[feat]]$sd
    }

    if (is.na(feat_sd) || feat_sd == 0) {
      warning("Feature has zero/undefined standard deviation: ", feat)
      feat_sd <- 1
    }

    computed_params[[feat]] <- list(mean = feat_mean, sd = feat_sd)
    data[[paste0(feat, suffix)]] <- (data[[feat]] - feat_mean) / feat_sd
  }

  list(data = data, params = computed_params)
}

features_to_scale <- c(
  "bmi", "baseline_charlson", "mortality_rsi",
  "ccs_mort30rate", "ccs_complication_rate",
  "complication_rsi", "age", "hour"
)

train_scaled <- standardize_features(trainDat, features_to_scale)
trainDat     <- train_scaled$data

test_scaled  <- standardize_features(testDat, features_to_scale, params = train_scaled$params)
testDat      <- test_scaled$data

cat("Standardization complete.\n")
```

------------------------------------------------------------------------

## 7. Feature Engineering

Four new features are created to capture clinically meaningful risk patterns.

| Feature             | Description                              |
|---------------------|------------------------------------------|
| `comorbidity_count` | Sum of all baseline condition indicators |
| `obese`             | BMI ≥ 30                                 |
| `weekend_surgery`   | Surgery on Saturday or Sunday            |
| `night_surgery`     | Surgery at or before 7am or at/after 5pm |

```{r feature-engineering}
add_engineered_features <- function(data) {
  data %>%
    mutate(
      comorbidity_count = baseline_cancer + baseline_cvd + baseline_dementia +
        baseline_diabetes + baseline_digestive + baseline_osteoart +
        baseline_psych + baseline_pulmonary,
      obese          = as.numeric(bmi >= 30),
      weekend_surgery = as.numeric(dow == 0 | dow == 6),
      night_surgery   = as.numeric(hour >= 17 | hour <= 7)
    )
}

trainDat <- add_engineered_features(trainDat)
testDat  <- add_engineered_features(testDat)

# Final consistency checks
stopifnot("complication" %in% names(trainDat))
stopifnot("complication" %in% names(testDat))
stopifnot(all(trainDat$complication %in% c(0, 1)))
stopifnot(all(testDat$complication  %in% c(0, 1)))
stopifnot(isTRUE(all.equal(names(trainDat), names(testDat))))

cat("Feature engineering complete.\n")
```

```{r skim-train, results='markup'}
skimr::skim(trainDat)
```

```{r save-processed}
saveRDS(trainDat, file = paste0(params$my_path, "trainDat.rds"))
saveRDS(testDat,  file = paste0(params$my_path, "testDat.rds"))
cat("Processed datasets saved.\n")
```

------------------------------------------------------------------------

## 8. Learner Specification

Three predictor sets of increasing complexity are defined. Each is paired with one or more learners.

```{r predictor-sets}
outcomeName <- "complication"

predSet_bm <- c(
  "bmi", "age",
  "asa_status_1", "asa_status_2",
  "baseline_cancer", "baseline_cvd", "baseline_dementia",
  "baseline_diabetes", "baseline_digestive", "baseline_osteoart",
  "baseline_psych", "baseline_pulmonary", "baseline_charlson"
)

predSet_extended <- unique(c(
  predSet_bm,
  paste0("ahrq_ccs_", 0:20),
  "weekend_surgery", "gender", "ccs_complication_rate",
  "race_1", "race_2", "comorbidity_count"
))

predSet_kitchensink <- unique(c(
  predSet_extended,
  "complication_rsi", "mortality_rsi"
))

allPredsets <- list(
  predSet_bm          = predSet_bm,
  predSet_extended    = predSet_extended,
  predSet_kitchensink = predSet_kitchensink
)

specifiedLearners <- list(
  predSet_bm          = c("glm"),
  predSet_extended    = c("glm", "lasso", "random_forest"),
  predSet_kitchensink = c("lasso", "random_forest", "xgboost")
)

# Validation
missingNames <- setdiff(names(specifiedLearners), names(allPredsets))
if (length(missingNames) > 0) {
  stop("Learner names do not match predictor sets: ", paste(missingNames, collapse = ", "))
}

cat("Predictor set sizes:\n")
cat("  Benchmark:    ", length(predSet_bm),          "predictors\n")
cat("  Extended:     ", length(predSet_extended),     "predictors\n")
cat("  Kitchen sink: ", length(predSet_kitchensink),  "predictors\n")
```

```{r recipes-cv}
recipes_list <- lapply(names(allPredsets), function(name) {
  formula <- as.formula(
    paste(outcomeName, "~", paste(allPredsets[[name]], collapse = "+"))
  )
  recipe(formula, data = trainDat) %>%
    step_bin2factor(all_of(outcomeName))
})
names(recipes_list) <- names(allPredsets)
saveRDS(recipes_list, file = paste0(params$my_path, "recipes.rds"))

recipe_main <- recipe(as.formula(paste(outcomeName, "~ .")), data = trainDat) %>%
  step_bin2factor(all_of(outcomeName))
saveRDS(recipe_main, file = paste0(params$my_path, "recipe_main.rds"))

# Cross-validation and tuning parameters
cvFolds     <- 5
stratVar    <- "complication"
gridSize    <- 5
tuneMetrics <- "pr_auc"

cvParams <- list(
  cvFolds     = cvFolds,
  stratVar    = stratVar,
  gridSize    = gridSize,
  tuneMetrics = tuneMetrics
)

equityVars         <- c("gender", "race_0", "race_1", "race_2")
positiveProportion <- mean(trainDat[[outcomeName]] == 1)

learnerSpec <- list(
  my_path            = params$my_path,
  trainDataName      = "trainDat.rds",
  testDataName       = "testDat.rds",
  outcomeName        = outcomeName,
  allPredsets        = allPredsets,
  specifiedLearners  = specifiedLearners,
  cvParams           = cvParams,
  equityVars         = equityVars,
  positiveProportion = positiveProportion
)

saveRDS(learnerSpec, file = paste0(params$my_path, "learnerSpec.rds"))
cat("Learner specifications saved.\n")
```

------------------------------------------------------------------------

## 9. Model Training and Validation

```{r load-helpers}
source("/share/dsa435601s26/PA-Class/R/packages.R")
source("/share/dsa435601s26/PA-Class/R/learner_training.R")
source("/share/dsa435601s26/PA-Class/R/learner_training_helpers.R")
source("/share/dsa435601s26/PA-Class/R/equity_metrics.R")
source("/share/dsa435601s26/PA-Class/R/performance_metrics.R")
source("/share/dsa435601s26/PA-Class/R/plots_updated.R")
```

```{r parallel-setup}
if (requireNamespace("doParallel", quietly = TRUE)) {
  library(doParallel)
  n_cores <- max(1, parallel::detectCores() - 1)
  cl      <- makePSOCKcluster(n_cores)
  registerDoParallel(cl)
  cat("Parallel processing enabled with", n_cores, "cores\n")
}
```

```{r train-or-load}
if (isTRUE(params$load_run)) {
  learnersResults <- readRDS(paste0(params$my_path, "learnerResults.rds"))
  cat("Loaded existing learner results.\n")
} else {
  start_time      <- Sys.time()
  learnersResults <- train_learners(
    trainDat    = trainDat,
    learnerSpec = learnerSpec,
    recipes     = recipes_list,
    mainRecipes = recipe_main
  )
  end_time <- Sys.time()
  cat("Training completed in",
      round(difftime(end_time, start_time, units = "mins"), 1), "minutes\n")
  saveRDS(learnersResults, file = paste0(params$my_path, "learnerResults.rds"))
}

modelResults <- learnersResults$modelResults
predProbs    <- learnersResults$predProbs

cat("Learners trained:", length(table(predProbs$learnerName)), "\n")
```

### Validation Plots

```{r validation-plots}
plot_predicted_probabilities(predProbs)
plot_predicted_probabilities_split(predProbs, learnerSpec$outcomeName)
plot_roc_curve_with_reference(learnersResults$rocResults, referencePredSet = "predSet_bm")
plot_pr_curve_with_reference(learnersResults$prResults,  referencePredSet = "predSet_bm")
```

### Classification Metrics at Selected Thresholds

```{r threshold-metrics}
selectedLearners  <- c(
  "random_forest_predSet_extended",
  "random_forest_predSet_kitchensink",
  "xgboost_predSet_kitchensink"
)
selectedThresholds <- c(0.25, 0.35)

learnerMetricsThres <- calc_threshold_metrics(
  predProbs   = predProbs,
  outcomeName = learnerSpec$outcomeName,
  thresholds  = selectedThresholds
)

rateMetrics  <- c("tpr", "fpr", "tnr", "fnr", "ppv", "f1", "acc")
countMetrics <- c("tpc", "fpc", "tnc", "fnc")
allMetrics   <- c(rateMetrics, countMetrics)

plotData <- learnerMetricsThres %>%
  filter(learnerName %in% selectedLearners, threshold %in% selectedThresholds) %>%
  pivot_longer(cols = all_of(allMetrics), names_to = "metricType", values_to = "value")

outcomeDescription <- c("No surgical complications", "Surgical complications")

conf_mat_bar_plot(
  plotData           = plotData %>% filter(metricType %in% rateMetrics),
  metricType         = "rate",
  outcomeDescription = outcomeDescription,
  graphTitle         = "Classification Performance Metrics"
)
```

### Fairness Assessment — Validation Set

```{r fairness-validation}
for (equityVar in learnerSpec$equityVars) {
  if (equityVar %in% names(trainDat)) {
    predProbsWithGroup <- predProbs %>%
      bind_cols(trainDat %>% select(all_of(equityVar)))

    print(plot_pred_probs(
      predProbsWithGroup,
      learner   = "random_forest_predSet_extended",
      group_var = equityVar
    ))

    print(classification_metrics_by_group(
      predProbs   = predProbsWithGroup,
      outcomeName = learnerSpec$outcomeName,
      learnerName = "random_forest_predSet_extended",
      threshold   = selectedThresholds,
      groupVar    = equityVar
    ))
  }
}
```

### Final Validation Model Selection

The selected model is `random_forest_predSet_extended` at a threshold of 0.25. The lower threshold reflects the clinical priority of catching likely complications, where the cost of a false negative outweighs the cost of a false positive.

```{r final-selection}
finalLearner     <- "random_forest_predSet_extended"
finalThresholds  <- c(0.25)

cat("Selected learner:   ", finalLearner,    "\n")
cat("Selected threshold: ", finalThresholds, "\n")

modelResults %>%
  filter(learnerName == finalLearner) %>%
  select(.metric, mean, std_err) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "Validation metrics for selected learner")

learnerMetricsThres %>%
  filter(learnerName == finalLearner, threshold %in% finalThresholds) %>%
  select(threshold, all_of(allMetrics)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "Classification metrics at selected threshold")
```

```{r save-validation}
validateLearnersResults <- list(
  modelResults      = modelResults %>% filter(learnerName == finalLearner),
  predProbs         = predProbs    %>% filter(learnerName == finalLearner),
  modelFits         = learnersResults$modelFits[finalLearner],
  selectedThresholds = finalThresholds
)

saveRDS(validateLearnersResults, file = paste0(params$my_path, "validateLearnersResults.rds"))

# Stop parallel cluster if it was started
if (exists("cl")) parallel::stopCluster(cl)
```

------------------------------------------------------------------------

## 10. Final Test-Set Evaluation

```{r load-validation-results}
trainResults      <- readRDS(paste0(params$my_path, "validateLearnersResults.rds"))
trainPredProbs    <- trainResults$predProbs
outcomeName       <- learnerSpec$outcomeName
learnerName       <- trainResults$modelResults$learnerName[[1]]
selectedThreshold <- trainResults$selectedThresholds
```

```{r predictor-check}
model_vars        <- colnames(trainResults$modelFits[[learnerName]]$pre$mold$predictors)
missing_in_test   <- setdiff(model_vars, names(testDat))

if (length(missing_in_test) == 0) {
  cat("All required predictors are present in the test dataset.\n")
} else {
  stop("Missing predictors in test set: ", paste(missing_in_test, collapse = ", "))
}
```

### Key Predictor Distribution: Training vs. Test

```{r distribution-check}
key_predictors <- intersect(
  c("age", "bmi", "asa_status_1", "asa_status_2",
    "ccs_complication_rate", "comorbidity_count"),
  model_vars
)

for (pred in key_predictors) {
  cat("Variable:", pred, "\n")
  cat("  Training — mean:", round(mean(trainDat[[pred]]), 3),
      "| sd:", round(sd(trainDat[[pred]]), 3), "\n")
  cat("  Test     — mean:", round(mean(testDat[[pred]]),  3),
      "| sd:", round(sd(testDat[[pred]]),  3), "\n\n")
}
```

```{r test-predictions}
testPredProbs <- apply_learner_to_test(
  modelFit    = trainResults$modelFits[[learnerName]],
  testDat     = testDat,
  learnerName = learnerName,
  outcomeName = outcomeName
)
```

### Discrimination Metrics

```{r test-discrimination}
testRocResults <- calc_roc_curve(testPredProbs, outcomeName = outcomeName)
testPrResults  <- calc_pr_curve(testPredProbs,  outcomeName = outcomeName)
testMetrics    <- calc_auc_metrics(testPredProbs, outcomeName = outcomeName)

plot_roc_curve_with_reference(
  testRocResults,
  referencePredSet = unique(testRocResults$predSet),
  title = "Test Set ROC Curve"
)
plot_pr_curve_with_reference(
  testPrResults,
  referencePredSet = unique(testPrResults$predSet),
  title = "Test Set Precision-Recall Curve"
)
```

### Validation vs. Test Performance

```{r val-vs-test}
comparison <- tibble(
  Metric       = c("ROC-AUC", "PR-AUC"),
  Validation   = c(
    round(trainResults$modelResults %>% filter(.metric == "roc_auc") %>% pull(mean), 3),
    round(trainResults$modelResults %>% filter(.metric == "pr_auc")  %>% pull(mean), 3)
  ),
  `Test Set`   = c(
    round(testMetrics$roc_auc, 3),
    round(testMetrics$pr_auc,  3)
  )
)

knitr::kable(comparison, caption = "Validation vs. Test Discrimination Metrics")
```

### Classification Metrics at Selected Threshold

```{r test-threshold-metrics}
testMetricsThres <- calc_threshold_metrics(
  predProbs   = testPredProbs,
  outcomeName = outcomeName,
  thresholds  = selectedThreshold
)

testMetricsThres %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "Test set classification metrics at threshold 0.25")

plotData <- testMetricsThres %>%
  filter(threshold %in% selectedThreshold) %>%
  pivot_longer(cols = all_of(allMetrics), names_to = "metricType", values_to = "value")

conf_mat_bar_plot(
  plotData           = plotData %>% filter(metricType %in% rateMetrics),
  metricType         = "rate",
  outcomeDescription = outcomeDescription,
  graphTitle         = "Test Set Classification Performance"
)
```

### Fairness Assessment — Test Set

```{r fairness-test}
allEquityMetrics <- list()
allEquityTests   <- list()

for (equityVar in learnerSpec$equityVars) {
  if (equityVar %in% names(testDat)) {

    if (!(equityVar %in% names(testPredProbs))) {
      testPredProbs <- testPredProbs %>%
        bind_cols(testDat %>% select(all_of(equityVar)))
    }

    cat("Assessing fairness for:", equityVar, "\n")

    print(plot_pred_probs(
      testPredProbs,
      learner   = unique(testPredProbs$learnerName),
      group_var = equityVar
    ))

    allEquityMetrics[[equityVar]] <- classification_metrics_by_group(
      predProbs   = testPredProbs,
      outcomeName = outcomeName,
      learnerName = learnerName,
      threshold   = selectedThreshold,
      groupVar    = equityVar
    )

    allEquityTests[[equityVar]] <- test_group_disparities(
      predProbs   = testPredProbs,
      outcomeName = outcomeName,
      learnerName = learnerName,
      threshold   = selectedThreshold,
      groupVar    = equityVar
    )
  }
}

bind_rows(allEquityMetrics) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "Fairness metrics by demographic group")

bind_rows(allEquityTests) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "Statistical tests for group disparities")
```

```{r save-final}
finalResults <- list(
  learnerName       = learnerName,
  selectedThreshold = selectedThreshold,
  testPredProbs     = testPredProbs,
  testMetrics       = testMetrics,
  testMetricsThres  = testMetricsThres,
  equityMetrics     = bind_rows(allEquityMetrics),
  equityTests       = bind_rows(allEquityTests)
)

saveRDS(finalResults, file = paste0(params$my_path, "finalTestResults.rds"))
cat("Final test results saved.\n")
```
